## R Markdown

---
title: "5-hmC non-zero probe count 061018"
author: "Joe Kochmanski"
output:
  html_document:
    code_folding: hide
    fig_height: 2.5
    fig_width: 3.5
    highlight: tango
    toc: yes
    toc_float: yes
---
## Project summary

This script intersects and counts significant probes and probes from filtered oxBS-MLE beta data.

1. Import the raw beta values from oxBS-MLE output and the output from glmmTMB and gamlss for EPIC data (significant probes).
2. Filter raw beta values down to only probes with non-zero values
3. Intersect significant probes with filtered beta probes
4. Save and explore remaining probe list
5. Filter raw beta values down to probes with at least three non-zero values in one disease group.
6. Intersect significant probes with filtered beta probes
7. Save and explore remaining probe list

## Load required packages        
```{r setup, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE, results='hide'}
library(dplyr)
library(knitr)
library(ggplot2)
```

## 1. Load data
```{r Load data, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
load("./oxBSMLE_betas.RData")
load("./5hmC/fivehmc_glialadjusted_nopmi_cingulate.RData")
load("./5hmC/fivehmc_glialadjusted_nopmi_parietal.RData")
```

## 2. Filter data

```{r Create data.frames of the 5-hmC beta data, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
hydroxy_mC_C = as.data.frame(hydroxymethylcyto.c) #766212 probes
hydroxy_mC_P = as.data.frame(hydroxymethylcyto.p) #774831 probes
```

```{r Go through each row and determine if a beta value is zero, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
row_sub = apply(hydroxy_mC_C, 1, function(row) all(row !=0 ))
row_sub1 = apply(hydroxy_mC_P, 1, function(row) all(row !=0 ))
```

```{r Subset the beta data to remove rows with zeros, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
hydroxy_C_nonzero <- hydroxy_mC_C[row_sub,] #Removes all rows w/ zero as a value.
hydroxy_P_nonzero <- hydroxy_mC_P[row_sub1,] #Removes all rows w/ zero as a value.
```

## 3. Intersect and Count Non-zero Probes

### Intersect data to generate filtered data.frame
```{r Merge data.frames for filtered 5hmC beta data and significant probes using "Probe" column, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
hmc.cing.late.nonzero.merged <- merge(hydroxy_C_nonzero, hydroxymethylcyto.all.cing.late.nopmi, by=c("Probe"))
hmc.cing.mid.nonzero.merged <- merge(hydroxy_C_nonzero, hydroxymethylcyto.all.cing.mid.nopmi, by=c("Probe"))
hmc.pari.late.nonzero.merged <- merge(hydroxy_P_nonzero, hydroxymethylcyto.all.pari.late.nopmi, by=c("Probe"))
hmc.pari.mid.nonzero.merged <- merge(hydroxy_P_nonzero, hydroxymethylcyto.all.pari.mid.nopmi, by=c("Probe"))
```

### Count intersecting non-zero probes by stage
```{r Separate probe modifications in mid stage, late stage, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
library(dplyr)
c.mid.hmc.nonzero <- dplyr::intersect(hydroxy_C_nonzero$Probe, hydroxymethylcyto.all.cing.mid.nopmi$Probe)
p.mid.hmc.nonzero <- dplyr::intersect(hydroxy_P_nonzero$Probe, hydroxymethylcyto.all.pari.mid.nopmi$Probe)

c.late.hmc.nonzero <- dplyr::intersect(hydroxy_C_nonzero$Probe, hydroxymethylcyto.all.cing.late.nopmi$Probe)
p.late.hmc.nonzero <- dplyr::intersect(hydroxy_P_nonzero$Probe, hydroxymethylcyto.all.pari.late.nopmi$Probe)
```

```{r Count non-zero significant probes by disease stage, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
c.mid.hmc.counts <- as.numeric(c(length(c.mid.hmc.nonzero)))
c.late.hmc.counts <- as.numeric(c(length(c.late.hmc.nonzero)))
p.mid.hmc.counts <- as.numeric(c(length(p.mid.hmc.nonzero)))
p.late.hmc.counts <- as.numeric(c(length(p.late.hmc.nonzero)))
probe_counts <- data.frame(c.mid.hmc.counts,c.late.hmc.counts,p.mid.hmc.counts,p.late.hmc.counts)
names(probe_counts) <- c("Cingulate-Mid","Cingulate-Late","Parietal-Mid","Parietal-Late")
row.names(probe_counts) <- c("5hmC non-zero probes")                         
kable(probe_counts) 
```

##4. Save and Explore Data

```{r Save probes sorted by stage as RData, echo=FALSE, warning=FALSE, message=FALSE, cache=FALSE}
save(hmc.cing.late.nonzero.merged,hmc.cing.mid.nonzero.merged,hmc.pari.late.nonzero.merged,hmc.pari.mid.nonzero.merged, file = "./sig.hmc.nonzero.stage.RData")
```

```{r Save probes sorted by stage as .csv files, echo=FALSE, warning=FALSE, message=FALSE, cache=FALSE}
write.csv(hmc.cing.late.nonzero.merged, "./P1_Cing_Late_Nonzero_Merged_SigProbes_061018.csv")
write.csv(hmc.cing.mid.nonzero.merged, "./P1_Cing_Mid_Nonzero_Merged_SigProbes_061018.csv")
write.csv(hmc.pari.late.nonzero.merged, "./P1_Pari_Late_Nonzero_Merged_SigProbes_061018.csv")
write.csv(hmc.pari.mid.nonzero.merged, "./P1_Pari_Mid_Nonzero_Merged_SigProbes_061018.csv")
```


##5. Filter oxBS mle estimates of 5hmc to only probes that had 3 non-zero values in at least one disease stage.

```{r Create data.frames of the 5-hmC beta data, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
#To examine number of non-zero probes in each disease stage, split 5-hmC #data.frames into 5-hmC beta data for each disease stage.
hydroxy_mC_C_1 = hydroxy_mC_C[1:3] #766212 probes
hydroxy_mC_C_2 = hydroxy_mC_C[4:6] 
hydroxy_mC_C_3 = hydroxy_mC_C[7:9] 

hydroxy_mC_P_1 = hydroxy_mC_P[1:3] #774831 probes
hydroxy_mC_P_2 = hydroxy_mC_P[4:6] 
hydroxy_mC_P_3 = hydroxy_mC_P[7:9] 
```

```{r Go through each row and determine if a value is zero, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
row_sub_C_1 = apply(hydroxy_mC_C_1, 1, function(row) all(row !=0 ))
row_sub_P_1 = apply(hydroxy_mC_P_1, 1, function(row) all(row !=0 ))

row_sub_C_2 = apply(hydroxy_mC_C_2, 1, function(row) all(row !=0 ))
row_sub_P_2 = apply(hydroxy_mC_P_2, 1, function(row) all(row !=0 ))

row_sub_C_3 = apply(hydroxy_mC_C_3, 1, function(row) all(row !=0 ))
row_sub_P_3 = apply(hydroxy_mC_P_3, 1, function(row) all(row !=0 ))
```

```{r Subset the data to remove rows with zeros, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}

hydroxy_C_1_nonzero <- hydroxy_mC_C_1[row_sub_C_1,] #Removes all rows w/ zero as a value.
#613543 probes
hydroxy_C_2_nonzero <- hydroxy_mC_C_2[row_sub_C_2,] #Removes all rows w/ zero as a value.
#616237 probes
hydroxy_C_3_nonzero <- hydroxy_mC_C_3[row_sub_C_3,] #Removes all rows w/ zero as a value.
#568130 probes

hydroxy_P_1_nonzero <- hydroxy_mC_P_1[row_sub_P_1,] #Removes all rows w/ zero as a value.
#622871 probes
hydroxy_P_2_nonzero <- hydroxy_mC_P_2[row_sub_P_2,] #Removes all rows w/ zero as a value.
#595627 probes
hydroxy_P_3_nonzero <- hydroxy_mC_P_3[row_sub_P_3,] #Removes all rows w/ zero as a value.
#627211 probes
```

```{r Move 5-hmC beta data probe row names to new column for matching w/ significant probes, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}

hydroxy_P_1_nonzero$Probe <- rownames(hydroxy_P_1_nonzero)
hydroxy_P_2_nonzero$Probe <- rownames(hydroxy_P_2_nonzero)
hydroxy_P_3_nonzero$Probe <- rownames(hydroxy_P_3_nonzero)
#Note: "Probe" matches column name in sig probes outputs
hydroxy_C_1_nonzero$Probe <- rownames(hydroxy_C_1_nonzero) 
hydroxy_C_2_nonzero$Probe <- rownames(hydroxy_C_2_nonzero) 
hydroxy_C_3_nonzero$Probe <- rownames(hydroxy_C_3_nonzero) 
```

```{r Merge filtered non-zero probes across disease stages for cing and pari, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
#NOTE: all=TRUE keeps all unique rows after merging, eventually giving a complete set of probes with at least one disease stage with 3 non-zero values.
hydroxy_P_nonzero_merged <- merge(hydroxy_P_1_nonzero, hydroxy_P_2_nonzero, all=TRUE)
hydroxy_P_all_nonzero_merged <- merge(hydroxy_P_nonzero_merged, hydroxy_P_3_nonzero, all=TRUE)
#696124 probes

hydroxy_C_nonzero_merged <- merge(hydroxy_C_1_nonzero, hydroxy_C_2_nonzero, all=TRUE)
hydroxy_C_all_nonzero_merged <- merge(hydroxy_C_nonzero_merged, hydroxy_C_3_nonzero, all=TRUE)
#680709 probes
```

```{r Merge data.frames for filtered 5hmC beta data and significant probes using "Probe" column, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}

hmc.cing.late.3nonzero.merged <- merge(hydroxy_C_all_nonzero_merged, hydroxymethylcyto.all.cing.late.nopmi, by=c("Probe"))
hmc.cing.mid.3nonzero.merged <- merge(hydroxy_C_all_nonzero_merged, hydroxymethylcyto.all.cing.mid.nopmi, by=c("Probe"))
hmc.pari.late.3nonzero.merged <- merge(hydroxy_P_all_nonzero_merged, hydroxymethylcyto.all.pari.late.nopmi, by=c("Probe"))
hmc.pari.mid.3nonzero.merged <- merge(hydroxy_P_all_nonzero_merged, hydroxymethylcyto.all.pari.mid.nopmi, by=c("Probe"))
```

##4. Save and Explore Data

```{r Merge data.frames for filtered 5hmC beta data and significant probes using "Probe" column, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
#Save merged data.frames as Rdata file:
save(hmc.cing.late.3nonzero.merged,hmc.cing.mid.3nonzero.merged,hmc.pari.late.3nonzero.merged,hmc.pari.mid.3nonzero.merged, file = "./sig.hmc.3nonzero.in.1stage.RData")

#Write .csv files of each separate merged data.frame
write.csv(hmc.cing.late.3nonzero.merged, "./P1_Cing_Late_3Nonzero_in_OneStage_SigProbes_061018.csv")
write.csv(hmc.cing.mid.3nonzero.merged, "./P1_Cing_Mid_3Nonzero_in_OneStage_SigProbes_061018.csv")
write.csv(hmc.pari.late.3nonzero.merged, "./P1_Pari_Late_3Nonzero_in_OneStage_SigProbes_061018.csv")
write.csv(hmc.pari.mid.3nonzero.merged, "./P1_Pari_Mid_3Nonzero_in_OneStage_SigProbes_061018.csv")
```

```{r Count probes that intersect between the significant probes and filtered beta data, echo=TRUE, warning=FALSE, message=FALSE, cache=FALSE}
library(dplyr)
c.mid.hmc.3nonzero <- dplyr::intersect(hydroxy_C_all_nonzero_merged$Probe, hydroxymethylcyto.all.cing.mid.nopmi$Probe)
p.mid.hmc.3nonzero <- dplyr::intersect(hydroxy_P_all_nonzero_merged$Probe, hydroxymethylcyto.all.pari.mid.nopmi$Probe)

c.late.hmc.3nonzero <- dplyr::intersect(hydroxy_C_all_nonzero_merged$Probe, hydroxymethylcyto.all.cing.late.nopmi$Probe)
p.late.hmc.3nonzero <- dplyr::intersect(hydroxy_P_all_nonzero_merged$Probe, hydroxymethylcyto.all.pari.late.nopmi$Probe)


c.mid.hmc.counts <- as.numeric(c(length(c.mid.hmc.3nonzero)))
c.late.hmc.counts <- as.numeric(c(length(c.late.hmc.3nonzero)))
p.mid.hmc.counts <- as.numeric(c(length(p.mid.hmc.3nonzero)))
p.late.hmc.counts <- as.numeric(c(length(p.late.hmc.3nonzero)))
probe_counts <- data.frame(c.mid.hmc.counts,c.late.hmc.counts,p.mid.hmc.counts,p.late.hmc.counts)
names(probe_counts) <- c("Cingulate-Mid","Cingulate-Late","Parietal-Mid","Parietal-Late")
row.names(probe_counts) <- c("5hmC non-zero (at least 3 in one stage) probes")
kable(probe_counts) #Creates table for Rmarkdown output -- pretty nifty!
```
